---
title: "Assignment 2"
author: "Shivam Shekher"
date: "2025-09-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# k-NN for classification
#Loading required libraries
```{r}
#install.packages(c("caret", "class", "gmodels", "dplyr", "ISLR"))
library(caret)
library(class)
library(gmodels)
library(dplyr)
library(ISLR) 
set.seed(123)
```



#Loading dataset
```{r}
# Load dataset
df <- read.csv("UniversalBank.csv")

# Drop ID and ZIP
df <- df %>% select(-ID, -ZIP.Code)

```

#Convert categorical variable to dummy variables using caret
```{r}
# Convert Personal.Loan to factor
df$Personal.Loan <- factor(df$Personal.Loan, levels = c(0,1))

# One-hot encode Education (3 levels)
df$Education <- factor(df$Education, levels = c(1,2,3))
dummies <- model.matrix(~ Education - 1, data=df)
colnames(dummies) <- c("Education_1","Education_2","Education_3")
df <- cbind(df %>% select(-Education), dummies)


```

#Dividing data into training and validation set
```{r}
# Partition 60% train / 40% validation
trainIndex <- createDataPartition(df$Personal.Loan, p=0.6, list=FALSE)
train <- df[trainIndex, ]
valid <- df[-trainIndex, ]

# Normalize predictors
preProc <- preProcess(train %>% select(-Personal.Loan), method=c("range"))
trainX <- predict(preProc, train %>% select(-Personal.Loan))
validX <- predict(preProc, valid %>% select(-Personal.Loan))

trainY <- train$Personal.Loan
validY <- valid$Personal.Loan
```

#Preparing Customer Data
```{r}
# Create new customer
new_customer <- data.frame(
  Age=40, Experience=10, Income=84, Family=2, CCAvg=2,
  Mortgage=0, Securities.Account=0, CD.Account=0,
  Online=1, CreditCard=1,
  Education_1=0, Education_2=1, Education_3=0
)

# Matching the variable order
new_customer <- new_customer[, names(trainX)]

# Normalize using training preproc before knn prediction
new_customerX <- predict(preProc, new_customer)

# k=1 prediction
knn1_pred <- knn(train=trainX, test=new_customerX, cl=trainY, k=1, prob=TRUE)
knn1_pred


#How would this customer be classified?

#The customer in this case would be classified according to the nearest neighbor of the customer with k=1. Due to the structure of the dataset, the classification is likely to be 0 (no loan acceptance) but this result is not stable and can change depending on the impact of the particular neighbor in the training partition.


```

#Choosing k to Balance Overfitting and Underfitting
```{r}

# Use caret::train to tune k
ctrl <- trainControl(method="cv", number=5) # 5-fold CV on training
knn_tune <- train(
  x = trainX, y = trainY,
  method = "knn",
  tuneGrid = data.frame(k=seq(1,25,2)),
  trControl = ctrl,
  preProcess = c("center","scale") # caret will scale internally too
)
knn_tune
plot(knn_tune)
best_k <- knn_tune$bestTune$k
best_k

#What is a choice of k that balances between overfitting and ignoring predictors?

#A value of k usually chosen between 5 and 15 is usually a trade off between overfitting (very small k) and missing information on predictors (very large k). In reality, we choose the very k based on the results of models on the validation set. This makes the model responsive to significant customer variations and not too much affected by random noise.

```

#Displaying confusion matrix for validation data using the best k.

```{r}

# Predicting validation set
best_pred_val <- knn(
  train = trainX, test = validX, cl = trainY,
  k = best_k
)
cm_val <- confusionMatrix(best_pred_val, validY, positive="1")
cm_val


```

#Classify Given Customer with Best k
```{r}
#Classifying the customer:(Age = 40, Experience = 10, Income = 84, Family = 2, CCAvg = 2, Education_1 = 0, Education_2 = 1, Education_3 = 0, Mortgage = 0, Securities Account = 0, CD Account = 0, Online = 1 and Credit Card = 1) using the best k.

knn_best_pred_customer <- knn(
  train=trainX, test=new_customerX, cl=trainY, k=best_k, prob=TRUE
)
knn_best_pred_customer


```
#Repartitioning into 50% train, 30% validation, 20% test. Compare confusion matrices.

```{r}

# Split data 50/30/20
set.seed(123)
trainIndex50 <- createDataPartition(df$Personal.Loan, p=0.5, list=FALSE)
train50 <- df[trainIndex50, ]
temp <- df[-trainIndex50, ]
valIndex30 <- createDataPartition(temp$Personal.Loan, p=0.6, list=FALSE) 
val30 <- temp[valIndex30, ]
test20 <- temp[-valIndex30, ]

# Normalize using 50% training set
preProc2 <- preProcess(train50 %>% select(-Personal.Loan), method=c("range"))
train50X <- predict(preProc2, train50 %>% select(-Personal.Loan))
val30X <- predict(preProc2, val30 %>% select(-Personal.Loan))
test20X <- predict(preProc2, test20 %>% select(-Personal.Loan))

train50Y <- train50$Personal.Loan
val30Y <- val30$Personal.Loan
test20Y <- test20$Personal.Loan

# Train confusion (LOOCV to avoid self-predict bias)
train_pred <- knn.cv(train50X, cl=train50Y, k=best_k)
cm_train <- confusionMatrix(train_pred, train50Y, positive="1")

# Validation 30%
val_pred <- knn(train=train50X, test=val30X, cl=train50Y, k=best_k)
cm_val30 <- confusionMatrix(val_pred, val30Y, positive="1")

# Test 20%
test_pred <- knn(train=train50X, test=test20X, cl=train50Y, k=best_k)
cm_test20 <- confusionMatrix(test_pred, test20Y, positive="1")

cm_train
cm_val30
cm_test20
#The 50/30/20 split confusion matrix analysis.The accuracy of the three sets (train: 95.1, val: 95.7 and test: 94.9) is very similar. Sensitivity on any of the splits is moderate (~0.55 to 0.63).This consistency demonstrates that the model is generalizable: There is no significant overfitting or underfitting.The minor decrease in the accuracy of the validation and test is caused by the unseen data. Generally, there is no dramatic change in performance and the selected k is good in terms of balance of new customers.
```









